A Detailed Demographic Analysis of 2020 U.S. Presidential Election Through Multilevel Regression and Post-Stratification

### 2024-03-16 paper/paper.qmd
review

---
title: "A Detailed Analysis of 2020 U.S. Presidential Election: A Demographic Analysis Through Multilevel Regression and Post-Stratification"
subtitle: "The Impact of Gender, Educatoin, Geographic location on Voter Preferences"
author: 
  - Jeongwoo Kim
  - Jiwon Choi
thanks: "Code and data are available at: https://github.com/Kjeongwoo99/2020_US_Election_Forecast"
date: today
date-format: long
abstract: "In this study, we forecasted the 2020 U.S. Presidential Election outcomes using a sophisticated statistical technique called multilevel regression with post-stratification (MRP), applied to a large-scale national survey dataset and census data. Our findings reveal how demographic factors such as gender, education, and geographic location influence voting behavior, with particular emphasis on voter preferences for Donald Trump. The analysis highlights significant variations in voter support across gender, states and educational backgrounds and how they shape electoral outcomes. This research contributes to our understanding of the American electoral landscape, demonstrating the critical role of demographic diversity in determining the direction of political preferences and election results."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The analysis reveals critical insights into how education, race, gender and age influence voting behavior. Contrary to conventional wisdom and past electoral analyses, our findings indicate a divergent pattern in voting behavior among different educational groups. The shift in party allegiance among non-college voters from Republicans to Democrats and the the dominance of popularity of the Republicans with 'Some college' and 'Postgrad' education levels challenge the stereotypical narratives that educational attainment leads to an increase in support for the Democrats. This group, especially highly educated 'Postgrad' group being the second highest supporters suggest a reevaluation of Democrats' strategies to appeal to educated voters concerned with economic policies and national security. However, according to our analysis, the Republicans fail to attract voters as highest proportion of support among the four educational groups remain under 40%. This almost certainly predicts that Joe Biden will become the president of the United States. This coincides with the election results where Biden won 55% of the votes from the college graduates while Trump gained 43% [@theWashingtonPost]. Our analysis of 37% of voters college degree and 37.5% of postgrad suggest are in line with the results. However, this model has failed to estimate the winning of Trump from less educated group since we estimated 39% from voters with 'some college degree' and 36.5% from voters with 'High school or less' whereas the post-election results show 50% for Trump from 'Some college or less' [@theWashingtonPost].

@@

@tbl-model-coefficients shows how males were more likely than females to have a favorable view of Trump, which correctly captured the tendency of voters by sex since 53% of men favored Trump whereas 42% of women favored Trump [@theWashingtonPost]. Gender does not act as a uniform lens through which policy positions are viewed, however, it is still an important issue especially regarding Trump in this election due to his words and actions that may viewed as sexist. Historically, women have been leaning more liberal than men. This gender gap has varied across different elections but has remained a consistent feature of the American political landscape since the 1980s [@mary]. While men's preferences of a smaller government remained stable, women's preference for a bigger government increased [@pewresearchcenter]. This shift contributed to a widening gap in presidential job approval with men remaining more supportive attitude of Trump. Solanas (2018) suggests that Trump's adminstrations increased investment in health, education, and security for girls, teenagers, migrant women, and support measures for pregnant women, as well as the promotion of women in STEM fields and efforts against gender violence. However, this is no secret that Trump is very conservative in treating gender equality. His openly sexist comments and behaviours presaged potential limitations on women's rights and freedoms, which may have contributed to the gender gap further [@elcano].

@@

From @fig-distribution-by-state, we can observe that Trump leads in mid-west of America, Montana (MT), Wyoming (WY), Mississippi (MS), Maine (ME), Missouri (MO), Kentucky (KY), and Utah (UT). Even the competing states like Louisiana (LA), New Mexico (NM), Tennessee (TN) are also mid-west part of the country, which are rural areas with low population density. This is correctly estimated since in the real world as Trump won in these states: MT, WY, MS, MA, MI, KY, UT. He also won in LA and TN, which we considered were competing states. We can go more in depth of this study with the benefit of hindsight. We estimated that the biggest victory for Trump would be Montana (MT) with just below 70%. In reality, he won by 59.6%, which is lower than we thought but still sit within our error values. The next state with the highest proportion of support for Trump was Wyoming (WY) from our model with 65%, which is close to the election poll result of 69%. We were very successful with Mississippi (MS) with an estimation of 57% and the poll result of 57.5%. Maine (ME) was another very close estimate with 52% and the poll result of 53.1%. 56.8% voted for Trump in Missouri (MO), which was slightly higher than our estimation of 51%. We estimated that only 26% of voters in Hawaii (HI) would vote for Trump with error margins of ±10%, which matched 34.3% for Trump in the election. 41.9% for Trump in Colorado (CO) lie within our estimate of 37% with ±5% margins.

There were some cases where our estimation was incorrect in terms of who won the state but the results lie within our error margins. For instance, we estimated Utah (UT) in favor of the Democrats with a slight margin of 51% for Biden but Biden only received 37.6% in the election. This still lies within our error values that we estimated to be from 43% to 58%. However, there were some errors in our estimation where our model failed to predict the Trump winning in the number of states in general. Our model predicted Louisiana (LA) for Biden with 51% but he only received 39.9% of the votes. This value is lower than our error margins. 62% of the residents in Alabama, a Republican stronghold in the last eleven presidential elections had voted for Trump while our model estimated 45% of votes for Trump. The republicans won in Florida (FL) with a slight margin of 51.2% but we predicted an estimation of 45% with ±2% margins. What we estimated to be the strongest Democrat state was North Dakota (ND) with an estimation of only 23% with ±12% margins but in reality 65.1% of the residents voted for Trump. In general, big cities and large suburban areas along the west coast and the east coast favored Biden whereas Trump supporters are mostly in the rural areas in the mid-south and mid-north states from the election results.

Our model was able to capture some challenges of long-held assumptions about electoral tendecy by educaiton. Biden improved his performance among suburban voters and White non-college voters, demonstrating a broadening appeal across different demographic groups, which is what our model suggested [@pewresearchcenter1]. Trump maintained strong support among rural voters and White evangelicals, which underscore the importance of educational divides, with significant differences in candidate support based on education levels [@pewresearchcenter1]. In addition, while Trump's stronghold among White men without a college degree loosened somewhat, he still won a significant majority of this demographic, which is not captured in our model [@pewresearchcenter1].

We were able to estimate the winning states for each candidate with some success. From the data and analysis outlined above, we believe that our model suggests that Biden would have won both the popular vote and the electoral votes and become the president of the United States. This was partially right in that he did win both but with a smaller margin that we had expected. In the end, he won 306 electoral votes and won 51.3% of the votes to become the 46th US president [@CNN]. however, + add why this happened

@@

One major limitation of this research is the reliance on the quality of the survey data collected in October 2019. Any inaccuracies  COVID-19 affected the collection of the survey data, so we had to use the 2019 data. The election was held in November 2020, so there exists a time gap of one year since the data collection and the election, which may have affected the voting as one year gives enough time for voters to change their mind. In addition, any inaccuracies or biases within the survey could significantly impact the model's forecasts, + data lost in the process.

Future research should explore models that can accommodate multiple electoral outcomes, including third-party voting and more variables if possible. Using more accurate or applicable real-time data sources could complement traditional survey methods and offer fresh insights into voter preferences. Societal norms and political landscapes change over time. Therefore, future research should focus on emerging social issues that may influence voter behavior. In the 2024 Presidential election, climate change, technology policy and social justice could play more significant roles in shaping electoral outcomes.

>>>>

---
title: "A Detailed Demographic Analysis of 2020 U.S. Presidential Election Through Multilevel Regression and Post-Stratification"
subtitle: "The Impact of Gender, Educatoin, Geographic location on Voter Preferences"
author: 
  - Jeongwoo Kim
  - Jiwon Choi
thanks: "Code and data are available at: https://github.com/Kjeongwoo99/2020_US_Election_Result_Analysis"
date: today
date-format: long
abstract: "In this study, we forecasted the 2020 U.S. Presidential Election outcomes using a sophisticated statistical technique called multilevel regression with post-stratification (MRP), applied to a large-scale national survey dataset and census data. Our findings reveal how demographic factors such as gender, education, and geographic location influence voting behavior, with particular emphasis on voter preferences for Donald Trump. The analysis highlights significant variations in voter support across gender, states and educational backgrounds and how they shape electoral outcomes. This research contributes to our understanding of the American electoral landscape, demonstrating the critical role of demographic diversity in determining the direction of political preferences and election results."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The analysis reveals critical insights into how education, race, gender, and age influence voting behavior. Contrary to conventional wisdom and past electoral analyses, our findings indicate a divergent pattern in voting behavior among different educational groups. The shift in party allegiance among non-college voters from Republicans to Democrats and the dominance of popularity of the Republicans with 'Some college' and 'Postgrad' education levels challenge the stereotypical narratives that educational attainment leads to an increase in support for the Democrats. This group, especially the highly educated 'Postgrad' group being the second highest supporters suggests a reevaluation of Democrats' strategies to appeal to educated voters concerned with economic policies and national security. However, according to our analysis, the Republicans failed to attract voters as the highest proportion of support among the four educational groups remains under 40%. This almost certainly predicts that Joe Biden will become the president of the United States. This coincides with the election results where Biden won 55% of the votes from the college graduates while Trump gained 43% [@theWashingtonPost]. Our analysis of 37% of voters with college degrees and 37.5% of postgrad suggests this is in line with the results. However, this model has failed to estimate the winning of Trump from the less educated group since we estimated 39% from voters with 'some college degree' and 36.5% from voters with 'High school or less' whereas the post-election results show 50% for Trump from 'Some college or less' [@theWashingtonPost].

@@

@tbl-model-coefficients shows how males were more likely than females to have a favorable view of Trump, which correctly captured the tendency of voters by sex since 53% of men favored Trump whereas 42% of women favored Trump [@theWashingtonPost]. Gender does not act as a uniform lens through which policy positions are viewed, however, it is still an important issue especially regarding Trump in this election due to his words and actions that may viewed as sexist. Historically, women have been leaning more liberal than men. This gender gap has varied across different elections but has remained a consistent feature of the American political landscape since the 1980s [@mary]. While men's preferences for a smaller government remained stable, women's preference for a bigger government increased [@pewresearchcenter]. This shift contributed to a widening gap in presidential job approval with men remaining more supportive attitude of Trump. Solanas suggests that Trump's administration increased investment in health, education, and security for girls, teenagers, and migrant women, and support measures for pregnant women, as well as the promotion of women in STEM fields and efforts against gender violence [@elcano]. However, this is no secret that Trump is very conservative in treating gender equality. His openly sexist comments and behaviours presaged potential limitations on women's rights and freedoms, which may have contributed to the gender gap further [@elcano].

@@

From @fig-distribution-by-state, we can observe that Trump leads in the mid-west of America, Montana (MT), Wyoming (WY), Mississippi (MS), Maine (ME), Missouri (MO), Kentucky (KY), and Utah (UT). Even the competing states like Louisiana (LA), New Mexico (NM), and Tennessee (TN) are also mid-west part of the country, which are rural areas with low population density. This is correctly estimated since in the real world where Trump won in these states: MT, WY, MS, MA, MI, KY, and UT. He also won in LA and TN, which we considered were competing states. We can go more in-depth into this study with the benefit of hindsight. We estimated that the biggest victory for Trump would be Montana (MT) with just below 70%. In reality, he won by 59.6%, which is lower than we thought but still sits within our error values. The next state with the highest proportion of support for Trump was Wyoming (WY) from our model with 65%, which is close to the election poll result of 69%. We were very successful with Mississippi (MS) with an estimation of 57% and the poll result of 57.5%. Maine (ME) was another very close estimate with 52% and the poll result of 53.1%. 56.8% voted for Trump in Missouri (MO), which was slightly higher than our estimation of 51%. We estimated that only 26% of voters in Hawaii (HI) would vote for Trump with error margins of ±10%, which matched 34.3% for Trump in the election. 41.9% for Trump in Colorado (CO) lie within our estimate of 37% with ±5% margins.

There were some cases where our estimation was incorrect in terms of who won the state but the results lie within our error margins. For instance, we estimated Utah (UT) in favor of the Democrats with a slight margin of 51% for Biden but Biden only received 37.6% in the election. This still lies within our error values which we estimated to be from 43% to 58%. However, there were some errors in our estimation where our model failed to predict the Trump winning in the number of states in general. Our model predicted Louisiana (LA) for Biden with 51% but he only received 39.9% of the votes. This value is lower than our error margins. 62% of the residents in Alabama, a Republican stronghold in the last eleven presidential elections had voted for Trump while our model estimated 45% of votes for Trump. The Republicans won in Florida (FL) with a slight margin of 51.2% but we predicted an estimation of 45% with ±2% margins. What we estimated to be the strongest Democrat state was North Dakota (ND) with an estimation of only 23% with ±12% margins but in reality, 65.1% of the residents voted for Trump. In general, big cities and large suburban areas along the west coast and the east coast favored Biden whereas Trump supporters are mostly in the rural areas in the mid-south and mid-north states from the election results.

Our model was able to capture some challenges of long-held assumptions about electoral tendency by education. Biden improved his performance among suburban voters and White non-college voters, demonstrating a broadening appeal across different demographic groups, which is what our model suggested [@pewresearchcenter1]. Trump maintained strong support among rural voters and White evangelicals, which emphasizes the importance of educational divides, with significant differences in candidate support based on education levels [@pewresearchcenter1]. In addition, while Trump's stronghold among White men without a college degree loosened somewhat, he still won a significant majority of this demographic, which is not captured in our model [@pewresearchcenter1].

We were able to estimate the winning states for each candidate with some success. From the data and analysis outlined above, we believe that our model suggests that Biden would have won both the popular vote and the electoral votes and become the president of the United States. This was partially right in that he did win both but with a smaller margin than we had expected. In the end, he won 306 electoral votes and won 51.3% of the votes to become the 46th US president [@CNN]. however, + add why this happened

@@

One major limitation of this research is the reliance on the quality of the survey data collected in October 2019. Any inaccuracies of COVID-19 affected the collection of the survey data, so we had to use the 2019 data. The election was held in November 2020, so there exists a time gap of one year between the data collection and the election, which may have affected the voting as one year gives enough time for voters to change their minds. In addition, any inaccuracies or biases within the survey could significantly impact the model's forecasts, especially, through the data cleaning process, which involved dropping observations with missing data. We reduced the survey dataset from 6,146 to 5,570 observations and the post-stratification dataset from 3,239,553 to 2,334,234 observations. This resulted in a loss of approximately 10% of the survey data and 30% of the post-stratification data. Such reductions could potentially impact the accuracy of our forecast results.

Future research should explore models that can accommodate multiple electoral outcomes, including third-party voting and more variables if possible. Using more accurate or applicable real-time data sources could complement traditional survey methods and offer fresh insights into voter preferences. Societal norms and political landscapes change over time. Therefore, future research should focus on emerging social issues that may influence voter behavior. In the 2024 Presidential election, climate change, technology policy, and social justice could play more significant roles in shaping electoral outcomes.
